# Results






```{r}
library(ggplot2)
library(forcats)
library(janitor)
library('openxlsx')
library(tidyr)
library(stringr)
library(httr)
library(jsonlite)
library(forcats)
library(sf)
library(spData)
library(purrr)
library(purrr)
library(ggpubr)
library(plyr)
library(dplyr)
```

```{r}

## hey we're working here 
pages <- seq(1,2)
us_eq_data = adply(pages, 1,
            function(page_no){
              res <- GET(paste0("https://www.ngdc.noaa.gov/hazel/hazard-service/api/v1/earthquakes?country=USA&page=",
                                page_no))
              data1 <- fromJSON(rawToChar(res$content))
              return(data1$items)
            }
            )
            
```

```{r}
date_sequence <- seq(as.Date("1900-01-01"), as.Date("1995-12-01"), length.out=50)
global_eq_data <- map2_dfr(date_sequence[1:length(date_sequence)-1], 
                           date_sequence[-1], 
                           .f=function(x,y){
                             print(paste0(x, " - ", y))
                             # tryCatch()
                             res <- GET(url=paste0("https://earthquake.usgs.gov/fdsnws/event/1/query?format=csv&starttime=",x,"&endtime=",y,"&minmagnitude=4"))
                             return(read.csv(text=rawToChar(res$content)))
                           })
```

```{r}
## state and earthquakes table
state_earthquakes <- read.csv("data/earthquakes_by_state.csv")
state_pop <- read.xlsx("data/pop_data.xlsx")
```


```{r}
# The third row has the column names


state_pop <- state_pop %>%
  row_to_names(row_number = 3) %>%
  clean_names()

colnames(state_pop)[1] = 'State'

state_pop <- state_pop %>% 
  slice(6:(nrow(state_pop)-5) ) %>% 
  mutate(State = substring(State, 2)) %>% 
  select(-2) %>% 
  mutate(across(.cols = 2:ncol(.), .fns = as.numeric)) %>% 
  tibble() 


```


```{r}
state_pop
```





```{r}
(fivenum(state_earthquakes$YEAR))
```

```{r}
# Pivot the table so we have one row per year and state

population_per_state <- state_pop %>%
  pivot_longer(cols = starts_with("x"),
               names_to = "year",
               values_to = "population") %>%
  arrange(State, year)

# Display the updated data frame
population_per_state
```

```{r}

# Remove the 'x' and '.'
population_per_state <- population_per_state %>%
  mutate(year = as.numeric(str_remove(year, "x")))

population_per_state
```

```{r}
state_earthquakes_selected <- state_earthquakes %>%
  select(NAME,MAGNITUDE)
```




```{r,fig.width=10, fig.height=5}
# Group by state, calculate the average magnitude across all years


# Assuming your data frame is named 'your_data'

# Drop rows with NA values
state_earthquakes_selected_cleaned <- na.omit(state_earthquakes_selected)

# Select one row per state, grouping by and getting the average
one_per_state  <- state_earthquakes_selected_cleaned %>% mutate(col_a = 1) %>% select(state = NAME, magnitude=MAGNITUDE) %>% group_by(state) %>% summarise(magnitude=mean(magnitude))



ggplot(one_per_state, aes(x = magnitude)) +
  geom_bar(stat = "identity", aes(y=fct_reorder(state, magnitude)),fill = "skyblue") +
  labs(title = "Average Magnitude Per State for 1638-1985", x =  "Average Magnitude", y = "State") +
  theme_minimal()

```
As we can see, the average magnitude of earth quakes is highest in Idaho This is interesting because it seems as though the highest magnitutde earthquakes occur on the west coast. Nevada, Utah, Washington are  also all on the west coast, which brings up the argument that states on the west coast are more impacted by Earthquakes, in terms of how strong they are. There could be an argument that these states have a higher population than others so we will verify that soon.

```{r}
state_earthquakes %>% mutate(DECADE_START = YEAR - YEAR %% 10) %>% 
  drop_na(MAGNITUDE) %>% 
  group_by(DECADE_START, YEAR, MONTH, DAY, STATE) %>% 
  summarise(MAGNITUDE=mean(MAGNITUDE)) %>% 
  ungroup() %>% 
  mutate(YEAR = as.factor(YEAR), DECADE_START = as.factor(DECADE_START)) %>% 
  ggplot() + 
  geom_boxplot(aes(y=MAGNITUDE, x=DECADE_START), varwidth = TRUE, outlier.size = 0.1, fill="red", alpha=0.5) + 
  theme_linedraw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


```{r,fig.width=15, fig.height=5}


state_earthquakes %>% mutate(DECADE_START = YEAR - YEAR %% 10) %>%
    drop_na(MAGNITUDE) %>% 
    group_by(YEAR, MONTH, DAY, NAME) %>% 
    summarise(MAGNITUDE=mean(MAGNITUDE)) %>% 
    ungroup() %>% 
    mutate(YEAR = as.factor(YEAR)) %>% 
    ggplot() + 
    geom_boxplot(aes(y=MAGNITUDE, x=reorder(NAME, -MAGNITUDE, length, decreasing=TRUE)), 
                 varwidth = TRUE, outlier.size = 0.1, fill="red", alpha=0.5) + 
    theme_linedraw() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```
The plot above Shows the distribution of earthquakes throughout the years. As we can see, the earthquakes initially all shared the same distribution with respect to magnitude, and eventually grew enormously in range and variance. This makes sense as the decades progressed the technology of detection did too and therefore the measurements became more accurate. Turning to the efigure on the right, we're able to see that California is the most variable state in terms of Magntitude of earthquakes. This makes sense because it is the largest state in America and thus naturally have the most variable earthquakes. The plot on the right also displays how the N.A (unkown state) has a large distribution of earthquakes. This is the result of joins on unkown states between tables.


```{r}
## use the population from the most recent year
newest_year_pop <- population_per_state %>% 
  filter(year == max(year)) %>%
  select(-year)  # Drop the 'year' column




newest_year_pop


```

```{r}
library(dplyr)


# Perform left join for population and state. 
population_state <- left_join(newest_year_pop, state_earthquakes_selected_cleaned, by = c("State" = "NAME"))
population_state
```




```{r}
## Cleaning the table

counts_and_average_mag <- population_state %>%
  group_by(State) %>%
  summarise(Count = n(), Average_Magnitude = mean(MAGNITUDE, na.rm = TRUE))



# counts_and_average_mag <- counts_and_average_mag[-1, ]

## remove rows with numbers
rows_with_numbers <- grep("\\d", counts_and_average_mag$State)

# Remove rows with numbers
if(length(rows_with_numbers)>0){
  counts_and_average_mag <- counts_and_average_mag[-rows_with_numbers, ]
}


# Identify rows with ":" in the "State" column
rows_with_colon <- grepl(":", counts_and_average_mag$State)

# Remove rows with ":"
counts_and_average_mag <- counts_and_average_mag[!rows_with_colon, ]
counts_and_average_mag


```





```{r,fig.width=50, fig.height=50}

# Install and load the plotly library if not already installed
# install.packages("plotly")
# Install and load the plotly library if not already installed
# install.packages("plotly")
library(plotly)

state_plot <- state_earthquakes %>% 
  mutate(DECADE_START = YEAR - YEAR %% 50) %>% 
  mutate(DECADE_START = as.factor(DECADE_START)) %>% 
  select(DECADE_START, YEAR, MONTH, NAME) %>% 
  left_join(newest_year_pop, by = c("NAME" = "State" )) %>% 
  group_by(DECADE_START, NAME) %>%
  summarise(Count = n()) %>% 
  drop_na(NAME) %>% 
  mutate(NAME = fct_rev(fct_reorder(NAME, Count, .fun = sum, .desc = TRUE))) %>% 
  ungroup() %>% 
  mutate(cg = cut(Count, 4, breaks = fivenum(Count), labels = c(1, 2, 3, 4))) %>% 
  mutate(cg = if_else(NAME == "California", "CA", cg)) %>% 
  arrange(desc(NAME)) %>% 
  ggplot() +
  geom_point(aes(x = Count, y = fct_reorder(NAME, Count), col = DECADE_START)) +
  labs(title = "Earthquake Count Per State for  1638-1985", x = "Earthquake Count", y = "State") +
  theme_linedraw() + 
  facet_wrap(~ cg)

# Convert ggplot to plotly and adjust width
state_plotly <- ggplotly(state_plot, width = 1000, height = 1000)  # Adjust the width as needed

# Display the interactive plot
state_plotly



```
As we can see California has the highest number of earthquakes across all Decades. Another interesting point was that all counts across almost all states in the decades before the 1950s were extremely low. This could be due to the fact that at the time states could not track earthquakes well and therefore did not detect many of them.

```{r}
counts_pop_ <- population_state %>%
  group_by(State) %>%
  summarise(Count = n(),Average_Magnitude = mean(MAGNITUDE, na.rm = FALSE),the_population = mean(population, na.rm = FALSE))

## remove rows with numbers
counts_pop_
```


```{r}



counts_pop_ <- population_state %>%
  group_by(State) %>%
  summarise(Count = n(), Average_Magnitude = mean(MAGNITUDE, na.rm = FALSE),the_population = mean(population, na.rm = FALSE))






counts_pop_$Count <- round(log(counts_pop_$Count))
counts_pop_$the_population <- log(counts_pop_$the_population) ## took log so we can see it 
counts_pop_
```

```{r}
counts_pop_
```



```{r,fig.width=10, fig.height=5}




counts_pop_ <- na.omit(counts_pop_)

# Create bins for Average_Magnitude
counts_pop_$bins <- cut(counts_pop_$Average_Magnitude, breaks = c(0, 4, 11), labels = c("0-4", "5-11"))

# Scatter plot with facets
ggplot(counts_pop_, aes(x = the_population, y = Count)) +
  geom_smooth(method = "lm", se = FALSE, col = "blue") +  # Add a line of best fit

  geom_point() +
  facet_wrap(~ bins, scales = "free") +
  labs(title = "Scatter Plot of Current Population vs Frequency of Earthquakes with Binned Average Magnitude for 1638-1985",
       x = "Population (LOG)", y = "Frequency of Earthquake (LOG)")

```

There is a Slight positive correlation between states with a higher population and the frequency of which they have earthquakes. This is true for both high and low magnitutde earthquakes. This result, furthermore, motivates the need for safety measures against earthquakes across populus areas. If they do not, then there can be severe consequences for areas that are dense, in the form of lives lost.  

```{r,fig.width=10, fig.height=5}
usgs_eq_list <- read.csv('data/query.csv') %>% 
  tibble() %>% 
  separate(time, into = 'date', sep="T", extra = 'drop', remove = FALSE) %>% 
  mutate(date=as.Date(date))

earthquakes_locations <- usgs_eq_list %>% 
  st_as_sf( coords=4:3, crs=4326)

usgs_fault_lines <- st_read("data/global_fault_lines/gem_active_faults.shp", quiet=TRUE)
usgs_fault_lines <- st_set_crs(usgs_fault_lines, 4326)

## Find which points fall over land
ii <- !is.na(as.numeric(st_intersects(earthquakes_locations, world)))

## Check that it worked
plot(st_geometry(world), main="Global Earthquakes from 1900 to 2023 (Magnitude 7+)")
plot(earthquakes_locations, col=sapply(ii, function(x){if(x){"black"}else{"blue"}}), pch=16, add=TRUE, cex=0.5)
plot(st_geometry(usgs_fault_lines), col= rgb(red = 1, green = 0, blue = 0, alpha = 0.1), add=TRUE)
legend(120, -65,
     legend=c("Land","Water"),
     col=c("black","blue"),
     cex=2.5, pch=16, horiz=FALSE)
```
As we can see by inspection, most earthquakes occur near the fault lines. This makes sense and is pretty expected behavioiur. This plot also highlights Where most earthquakes occur globally. It's pretty clear that most occur on the coasts (where fault lines are most common) and thereby most occur on islands (near singapore and the philippines).

```{r}
max(us_eq_data$year)
```



```{r}
library(ggplot2)

# Assuming your data frame is named 'earthquake_data'
# Create a line plot of earthquake counts grouped by month
ggplot(us_eq_data, aes(x = month)) +
  geom_line(stat = "count", color = "blue") +
  labs(title = "Earthquake Counts by Month from 1500 to 2023",
       x = "Month",
       y = "Count") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_minimal()

```
Earthquakes occur most in November! This is really interesting. We'll now check it over the years. 

```{r}
library(ggplot2)

library(dplyr)

# Assuming your data frame is named 'us_eq_data'

# Filter the data to include only years after 1700
filtered_data <- us_eq_data %>%
  filter(year > 1700)

# Create a histogram of earthquake counts over the years
histogram_plot <- ggplot(filtered_data, aes(x = year)) +
  geom_histogram(binwidth = 1, fill = "red", color = "black") +
  labs(title = "Earthquake Frequencies from 1700 to 2023",
       x = "Year",
       y = "Frequency") +
  theme_minimal()

# Display the histogram
print(histogram_plot)



```
This plot displays the distribution of earthquakes from 1700 to 2023. As you can see, overtime, the frequencies increased which makes sense as measurement and systemic capturing of earthquakes improved.


```{r}
mosaic_data <- mutate(usgs_eq_list, Ocean = ii)
mosaic_data$horizontal_encode <-  ifelse(is.na(mosaic_data$horizontalError), '0', 'More than 0')


mosaicplot(table(mosaic_data$Ocean, mosaic_data$horizontal_encode),
           main = "Mosaic Plot: Ocean vs. Horizontal Error Category", color = c("blue", "green"))

```

```{r}
mosaic_data
```




```{r}
pts_split <- split(earthquakes_locations, (as.numeric(rownames(pts))-1) %/% 50) 
dists <- pts_split %>% 
    lapply(function(x){
      # distance in (m)
        dist <- st_geometry(obj = usgs_fault_lines) %>%
            st_distance(y=x) %>% 
            apply(2, min)
        return(x %>% cbind(tibble(dist)))
    })
```

```{r}
joined_list <-  list()
for(i in 1:length(pts_split)){
  join_list <- pts_split[[i]] %>% cbind(tibble(distance_to_fault = dists[[i]]))
  joined_list[[i]] <- join_list
}

earthquakes_with_fault_distance <- bind_rows(joined_list)
```

```{r}
earthquakes_with_fault_distance %>% 
  ggplot() +
  geom_point(aes(x=distance_to_fault/1000, y=mag), alpha=0.3) +
  ggtitle("Distance To Fault vs Magnitude") + xlab("Distance to Nearest Fault (km)") +
  xlim(0,1000) +
  theme_classic()
```
